Flooding of the Okavango Delta influences Connectivity for Dispersing African Wild Dogs

### 2022-12-22 Chapter_8/04_Manuscript/Manuscript.tex
upd

We capitalize on a ``natural experimental setup'', the flood-pulse driven
environmental change across the Okavango Delta in northern Botswana, to
investigate the impact of changing environmental conditions on dispersal
patterns and connectivity of the endangered African wild dog (\textit{Lycaon
pictus}). For this, we simulate dispersal trajectories across the landscape
under two extreme environmental scenarios; one assuming a maximum flood extent,
one assuming a minimum flood extent.

During maximum flood, we observe a reduction in connectivity and an increase in
dispersal durations between distinct habitat patches. Notably, dispersal into
the center of the Okavango delta is reduced by 80\% during maximum flooding. At
minimum flooding, conversely, the delta reveals vital dispersal corridors and
increases chances of successful dispersal into neighboring habitat patches.
Concomittantly, dispersal durations to move between patches is reduced at low
flood levels.

Climate change is expected to critically impact the flood-pulsing cycles of the
Okavango delta and may therefore substantially alter landscape connectivity in
this area. Besides a better understanding of the conservation needs for the
African wild dog, our study also presents a first step towards incorporating
environmental changes due to seasonality or climate change in dispersal and
connectivity analyses. Finally, our analysis demonstrates the usefulness of
individual-based dispersal simulations as a pertinent conservation tool to study
impacts of environmental change while accounting for dispersal movements.

@@

\section{Introduction}
\subsection{Climate Change}
Climate change is expected to profoundly impact ecosystems worldwide, with
far-reaching consequences for species living therein (cite IPCC)
\citep{Ozgul.2010, Radchuk.2019}. However, empirical information on species' responses to climate change remains
scarce, especially for regions most vulnerable to altered conditions
\citep{Paniw.2021}.

\subsection{Dispersal}
The lack of knowledge in terms of species ability to cope with climate change is
further compounded by a lack of information on dispersal under changing
environmental conditions \citep{Travis.2013}. By bringing individuals away from
their natal location, dispersal strongly influences population dynamics
\citep{Clobert.2012}; it enables genetic exchange \citep{Frankham.2002,
Leigh.2012, Baguette.2013}, facilitates the colonization of empty habitats
\citep{Gustafson.1996, Hanski.1999b, MacArthur.2001}, and promotes reinforcement
of weakened and small subpopulations \citep{Brown.1977}. Dispersal also serves
as a means to track suitable habitat \citep{Raia.2012} and may thus serve as a
means to mediate the demographic consequences of environmental change
\citep{Hodgson.2009, Travis.2013}. Dispersal is also inextricably linked to the
concept of landscape connectivity, which is generally understood as the degree
by which the landscape facilitates or impedes movements among habitat patches
\citep{Tylor.1993}.

As previously highlighted by \cite{Travis.2013}, climate may affect dispersal
directly, by altering the propensity of individuals to disperse, or indirectly,
through changes in the biophysical environment. Here, we will focus on the
latter and study dispersal prospects under changing environmental conditions.

In spite of its importance for determining population dynamics and landscape
connectivity, our understanding of dispersal and how dispersal is affected by
changing environmental conditions is limited. This is owed to the difficulty of
obtaining data collected on dispersing animals at the appropriate temporal and
spatial scale \citep{Graves.2014, Vasudev.2015} and our inability to project
dispersal prospects under changing environmental conditions into the future.

The ability of species
to persist under these changing conditions depends on the amplitude and speed of
environmental change and species' responses to such changes (cite someone).

It is, in fact, presumed that humanity has already transgressed the planetary
boundary for climate change, therefore triggering irreversible feedbacks in the
biosphere \citep{Rockstrom.2009}.

Cite IPCC, Planetary boundaries, Tucker (step length). Phenological shifts. Cite
Robynne, Arpat, Megan, Dominik, Rabaiotti etc.

\subsection{Connectivity}
Dispersal is an important, if not the important, driver of landscape
connectivity and therefore of major interest to conservation authorities. It has
also been demonstrated that dispersal may lead to population declines in areas
where anthropogenic mortality is high and dispersal prospects low
\citep{Leigh.2012}.

\subsection{Okavango Delta and African Wild Dogs}
The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmetal change on species dispersal ability and connectivity in
large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.
While global temperatures are expected to increase by xx degrees until the end
of the 21st century, For example, \cite{Engelbrecht.2015} predicts an increase
between 4 and 6\degree C for regions in southern Africa. One species that
heavily relies on dispersal for long-term persistence and that inhabits such an
arid ecosystem in Southern Africa is the African wild dog (\textit{Lycaon
pictus}). The African wild dog is a lightly built canid lives in cohesive packs
comprising up to 20 individuals. Wild dogs are social breeders and the majority
of reproduction is monopolized by the dominant alpha couple who are supported by
subordinates in taking care of their pups. The species exhibits strong
inbreeding avoidance, hence juvenile individuals born into the pack disperse in
same-sex sibling coalitions after reaching sexual maturity \citep{McNutt.1996,
Behr.2020}. Notably, the African wild dog is the sole extant representative of
its genus and thus considered as high-priority species for conservation efforts
\citep{Leigh.2012}. However, fragmentation and destruction of their historic
habitat may have decreased dispersal success for this species and thus
jeopardizes connectivity \citep{Leigh.2012}. Since the impacts of climate change
are not uniformly distributed across the planet, but particularly pronounced at
high latitudes and arid ecosystems, the species may be seen as most vulnerable
to such ecosystem shifts.

\subsection{Our Study}
Here, we capitalize on a previously parametrized dispersal model and use
dispersal simulations to investigate connectivity patterns for African wild dogs
under two extreme scenarios; one assuming maximum flooding of the Okavango delta
and one assuming minimum flooding of the Okavango delta. Given that water poses
a barrier to dispersing individuals, we anticipate that dispersal prospects and
connectivity during maximum flood are low. During minimum flood, in contrast, we
expect to reveal the presence of several dispersal corridors that may be used to
freely and quickly move between remaining habitat patches.

@@

\section{Results}
\subsection{Heatmaps}
Heatmaps produced from simulated dispersal trajectories reveal that the OD acts
as a major dispersal barrier during periods of high flood, but reveals viable
dispersal corridors during periods of low flood (\Cref{Heatmaps}). During
minimum flood, the area north-west of Maun appears to serve as vital dispersal
habitat. The same area is entirely avoided during maximum flood. Besides
striking differences in connectivity for the close vicinity of the delta, the
remainder of the study area shows only marginal differences in connectivity
between the two scenarios. For instance, in both scenarios the area south of the
Linyanti swamp appears as frequently visited dispersal habitat. Additional
heatmaps highlighting differences in connectivity for each source area
separately are provided in Appendix SX.

@@

@@

@@

@@

@@

@@

>>>>

We capitalize on a ``natural experimental setup'', the flood-pulse driven
environmental change across the Okavango Delta in northern Botswana, to
investigate the impact of changing environmental conditions on dispersal
patterns and connectivity of the endangered African wild dog (\textit{Lycaon
pictus}). For this, we simulate dispersal trajectories across the landscape of
the Okavango Delta under two extreme environmental scenarios; one assuming a
maximum flood extent, one assuming a minimum flood extent.

During maximum flood, we observe a reduction in connectivity and an increase in
dispersal durations between distinct habitat patches. Notably, dispersal into
the center of the Okavango delta is reduced by 80\% during maximum flooding. At
minimum flooding, conversely, the delta reveals vital dispersal corridors and
increases chances of successful dispersal into neighboring habitat patches by xx
percent. Concomitantly, dispersal durations to move between patches is reduced,
on average by xx days,at low flood levels.

Climate change is expected to critically impact the flood-pulsing cycles of the
Okavango delta and may therefore substantially alter landscape connectivity in
this area. Our analysis suggests that anticipating such alterations in
environmental conditions is key when implementing conservation strategies.
Besides a better understanding of the conservation needs for the African wild
dog, our study also presents a first step towards incorporating environmental
changes due to seasonality or climate change in dispersal and connectivity
analyses and demonstrates the usefulness of individual-based dispersal
simulations as a pertinent conservation tool to study impacts of environmental
change on dispersal success.

@@

\section{Introduction}
Climate change is expected to profoundly impact ecosystems worldwide with
far-reaching consequences for species living therein \citep{Ozgul.2010,
Radchuk.2019, IPCC.2022}. By altering environmental conditions, climate change
affects animal behavior \citep{Fuller.2016}, resource availability
\citep{Durant.2007}, population dynamics \citep{Paniw.2021}, and the
distribution of wild living animal populations \citep{Thomas.2004,
Thuiller.2006}. One potential life-history pathway through which species may
mediate the effects of environmental change is dispersal \citep{Anderson.2012}.
By bringing individuals away from their natal location \citep{Clobert.2012},
dispersal equips species with a means to adapt to climate change by tracking
favorable habitat conditions \citep{Raia.2012} and to shift into a different
part of their fundamental niches \citep{Kokko.2022}. Dispersal furthermore
provides resilience against changing conditions by ensuring genetic diversity
\citep{Frankham.2002, Leigh.2012, Baguette.2013}, facilitating the colonization
empty habitats \citep{Gustafson.1996, Hanski.1999b, MacArthur.2001}, and
promoting the reinforcement of weakened and small subpopulations
\citep{Brown.1977}. As such, it is little surpsing that improving connectivity
is viewed as the number one strategy to combat the adverse effects of climate
change \citep{Heller.2009}.

While dispersal offers a means to offset the demographic consequences of climate
change \citep{Kokko.2006, Hodgson.2009, Travis.2013}, it itself is a function of
climatic and environmental conditions (e.g. \citep{Behr.2020}). The link between
dispersal and climaticor environmental can either be indirect, for example if
the propensity of individuals to disperse depends on environmental conditions,
or direct, when the biophysical environment through which dispersers move
affects dispersal prospects \citep{Travis.2013}. As the latter case indicates,
dispersal is also inextricably linked to the concept landscape connectivity
\citep{Baguette.2013}, which is understood as the degree by which the landscape
facilitates or impedes movements \citep{Taylor.1993}. Only if a sufficient
degree of landscape connectivity is maintained, successful dispersal of
individuals can be guaranteed \citep{Fahrig.2003}. Unfortunately, our
understanding how changing environmental conditions influence connectivity and
dispersal is limited \citep{Kokko.2006, Travis.2013}. Because collecting
information on dispersing animals is difficult and because it is difficult to
assess how changing environmental conditions affect connectivity...

The continued degradation and destruction habitats across the globe impedes
dispersal, promotes inbreeding and genetic drift and increases the
likelihood of local extirpation due to stochastic events \citep{Sawyer.2011}.
Combined, these factors limit species ability to adapt to changing environmental
conditions \citep{Fahrig.2003}.

\cite{Ashrafzadeh.2019} found that cimate change is likely to substantially
reduce connectivity for mountain newts in Iran. Similarly, \cite{Luo.2021}
predicted significant reductions in connectivity for the giant spiny frog in
South-East Asia.

However, failing to take into account potential changes in climatic and
environmental conditions imperils our ability to design cost-effective
conservation strategies.

Here, we used naturally occuring seasonal flood-cycles to approximate the likely
changes in connectivity that will arise under amplified climatic conditions due
to climate change. While a better understanding of the ODs flooding regime under global change is of major interest, reliable predictions remain challenging due to the complex interplay between climate,
anthropogenic water use, animal behavior, and topography of the region.

Climate change
could lead to the disruption of existing dispersal corridors and the emergence
of new dispersal barriers but may also provide possibilities to access
previously unsuitable dispersal routes. Designing and implementing effective
conservation strategies that aim to improve landscape connectivity, such as
habitat restoration or the creation of wildlife corridors, requires better
understanding of such dynamics.

Albeit the impacts of climate change on connectivity cannot be directly
estimated, one can study how changing environmental conditions due to
seasonality influence patterns of dispersal and connectivity.

Habitat degradation and fragmentation lead to isolated subpopulations that have
a high associated risk of extinction due to stochastic events and generally
decreased genetic variability \citep{Melbourne.2008}.

More recently, individual-based movement models (IBMMs) have been uprising and
are recommended for their ability to incorporate several crucial realities of
animal behavior, such as, for example, autocorrelation, that are by omitted by
necessity using resistance-based approaches \citep{Hofmann.2023,
Unnithan.2022a}. IBMMs allow much greater flexibility and are thus considered
powerful tools for examining connectivity under different landscape
configurations \citep{Littlefield.2019}.

Botswana is among the countries most heavily influenced by climate change .

For multispecies approaches to connectivity within the KAZA-TFCA, see
\cite{Brennan.2020} and \cite{Lines.2021}.

Talk about connectivity models
Talk about functional connectivity
Talk about habitat fragmentation and destruction

Osipova

Unfortunately, little is known about the specific responses of different species
to climate change, especially for regions most vulnerable to altered conditions
\citep{Paniw.2021}.

Only if there

The challenges associated with studying dispersing animals have thus far
impaired the collection of data of dispersers at the appropriate temporal and
spatial scale \citep{Graves.2014, Vasudev.2015} and weakened our ability to
project dispersal prospects under changing environmental conditions into the
future. Nevertheless, a more comprehensive understanding of dispersal under
changing environmental conditions is critical for designing effective
conservation strategies in the light of climate change. By predicting and
modeling dispersal patterns under different scenarios, we can better anticipate
the impacts of climate change on species and design effective conservation
strategies to protect them. This is especially important for endangered species
that are already at the verge of extinction.

The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmental change on species dispersal ability and connectivity in a
large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.
While global temperatures are expected to increase by xx degrees until the end
of the 21st century, For example, \cite{Engelbrecht.2015} predicts an increase
between 4 and 6\degree C for regions in southern Africa. One species that
heavily relies on dispersal for long-term persistence and that inhabits such an
arid ecosystem in Southern Africa is the African wild dog (\textit{Lycaon
pictus}). The African wild dog is a lightly built canid lives in cohesive packs
comprising up to 20 individuals. Wild dogs are social breeders and the majority
of reproduction is monopolized by the dominant alpha couple who are supported by
subordinates in taking care of their pups. The species exhibits strong
inbreeding avoidance, hence juvenile individuals born into the pack disperse in
same-sex sibling coalitions after reaching sexual maturity \citep{McNutt.1996,
Behr.2020}. Notably, the African wild dog is the sole extant representative of
its genus and thus considered as high-priority species for conservation efforts
\citep{Leigh.2012}. However, fragmentation and destruction of their historic
habitat may have decreased dispersal success for this species and thus
jeopardizes connectivity \citep{Leigh.2012}. Since the impacts of climate change
are not uniformly distributed across the planet, but particularly pronounced at
high latitudes and arid ecosystems, the species may be seen as most vulnerable
to such ecosystem shifts.

Talk about Zeller und Osipova

In a previous study, \citep{Hofmann.2022} have investigated landscape
connectivity in the same ecosystem, yet assuming a static landscape.

Here, we capitalize on a previously parameterized dispersal model and use
dispersal simulations to investigate dispersal success and connectivity patterns
for African wild dogs under two extreme scenarios; one assuming maximum flooding
of the Okavango delta and one assuming minimum flooding of the Okavango delta.
Given that water poses a barrier to dispersing individuals, we anticipated that
dispersal prospects and connectivity during maximum flood are low. During
minimum flood, in contrast, we expect to reveal the presence of several
dispersal corridors that may be used to freely and quickly move between
remaining habitat patches.

@@

\section{Results}
\subsection{General}
See \Cref{Emigration}
\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Emigration.png}
  \caption{Caption}
  \label{Emigration}
  \end{center}
\end{figure}

\subsection{Heatmaps}
Heatmaps produced from simulated dispersal trajectories reveal that the OD acts
as a major dispersal barrier during periods of high flood, but reveals viable
dispersal corridors during periods of low flood (\Cref{Heatmaps}). During
minimum flood, the area north-west of Maun appears to serve as vital dispersal
habitat. The same area is entirely avoided during maximum flood. Besides
striking differences in connectivity for the close vicinity of the delta, the
remainder of the study area shows only marginal differences in connectivity
between the two scenarios. For instance, in both scenarios the area south of the
Linyanti swamp appears as frequently visited dispersal habitat. Additional
heatmaps highlighting differences in connectivity for each source area
separately are provided in Appendix SX.

@@

Although dispersal is generally believed to be beneficial for the long term viability of species, it can also lead to the net-loss of individuals in areas where edge effects and anthropogenic mortality are high \citep{Leigh.2012}.

@@

Dispersal is an important, if not the most important, driver of landscape
connectivity and therefore of major interest to conservation authorities. It has
also been demonstrated that dispersal may lead to population declines in areas
where anthropogenic mortality is high and dispersal prospects low
\citep{Leigh.2012}.

Efforts aiming at restoring or improving landscape connectivity should consider
and take into account changes in landscape characteristics due to climate
change.

@@

Climate change will not only change species distribution but also community
composition \citep{Thuiller.2006}.

@@

For the discussion: Through changing environmental conditions, climate change
will also impact the distribution of various species with far-reaching
consequences on inter- and intra-specific competition (Abrahms). Due to a lack
of data, we omitted any inclusion of social factors in our dispersal model,
albeit they are known to be important determinants of dispersal (Armansin).
Studying how climate change influences species distribution and their
interactions, and ultimately how this alters dispersal, will be challenging but
necessary to more realistically simulate dispersal.

@@

Although the OD is arguably the main driver of seasonal change across the
studied ecosystem, there are several other factors that undergoe seasonal
change, including vegetation and the abundance and distribution of prey or
predators.

@@

Although validating the predictions resulting from our dispersal model is
challenging, the simulations enable us to generate several interesting
hypotheses...
